### 2021-08-30 submission/manuscript.tex
Add edits and response to reviewers for revision

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients were
separated into two groups based on stool consistency, there were similar
microbiota features between samples from CDI patients and \emph{C.
difficile} negative patients with diarrhea compared to non-diarrheal
samples that were \emph{C. difficile} negative (12). The similar
community features between CDI patients and patients with diarrhea
included low alpha diversity and only 6 bacterial taxa with higher
relative abundances in communities from CDI patients. These results led
to the hypothesis that bacterial communities from patients experiencing
diarrhea are susceptible to developing CDIs, regardless of how they
developed diarrhea.

Depending on the dose administered, osmotic laxatives can lead to
diarrhea and temporarily disrupt the human intestinal microbiota (13).
The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
in Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after challenge is unclear.

@@

Because there were differences in the mucosal microbiota, including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
histopathology (27). However, we found there was no difference in cecum
and colon scores between clindamycin and PEG-treated mice that were
challenged with \emph{C. difficile} at 4 days post-challenge (Fig. 3D),
the time point typically examined in \emph{C. difficile} 630 challenged
mice (28). We also looked at 6 days post-challenge because that was when
there was a large difference in \emph{C. difficile} density between PEG-
and clindamycin-treated mice (Fig. 1C). Although there was a slight
difference in the histopathology score of the colon between PEG and
clindamycin-treated mice, there was not a signifant difference in the
cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
the mucosal microbiota, the impact of \emph{C. difficile} challenge on
the cecum and colon was similar between PEG and clindamycin treated
mice.

@@

While the disruptive effect of antibiotics on \emph{C. difficile}
colonization resistance is well established, the extent to which other
drugs such as laxatives disrupt colonization resistance was unclear. By
following mice treated with an osmotic laxative over time, we found that
a 5-day PEG treatment before challenge resulted in prolonged \emph{C.
difficile} colonization, while a 1-day PEG treatment resulted in
transient colonization without the use of antibiotics. The differences
in \emph{C. difficile} colonization dynamics between the 5- and 1-day
PEG treated mice were associated with differences in the degree to which
treatments disrupted the microbiota. Additionally, the intestinal
communities of 5-day PEG treated mice did not regain colonization
resistance after a 10-day recovery period. In contrast to the other
5-day PEG treatment groups, \emph{C. difficile} was not immediately
detectable in the stools of most of the mice in the 10-day recovery
group. We also examined the impact of PEG treatment after \emph{C.
difficile} challenge. In opposition to the hypothesis suggested by the
literature, we found that PEG treatment prolonged colonization relative
to mice that only recieved clindamycin treatment. We identified patterns
in the relative abundances of \emph{Bacteroides},
\emph{Enterobacteriaceae}, \emph{Akkermansia},
\emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
associated with prolonged \emph{C. difficile} colonization (Fig. 8).
Overall, our results demonstrated that osmotic laxative treatment alone
rendered mice susceptible to \emph{C. difficile} colonization and the
duration of colonization depended on the length of PEG treatment and
whether treatment was administered before or after challenge.

In addition to altering composition, laxative treatment may alter
microbiota-produced metabolites. A previous study demonstrated that a
5-day treatment of 10\% PEG depleted acetate and butyrate and increased
succinate compared to untreated mice (15). While we did not perform
metabolomic analysis, we did see bacteria known to produce beneficial
metabolites were depleted in mice that cleared \emph{C. difficile}
compared to mice with prolonged colonization. For example,
\emph{Oscillibacter valericigenes} can produce the SCFA valerate (30),
and separate studies demonstrated valerate is depleted after clindamycin
treatment and inhibited \emph{C. difficile} growth \emph{in vitro} and
in C57BL/6 mice (31, 32). Similarly, \emph{Acetatifactor} can produce
acetate and butyrate (33), SCFAs that are decreased in mice with
prolonged \emph{C. difficile} infection after antibiotic treatment (34).
Thus protective bacteria and their metabolites could be depleted by
osmotive laxative treatment depending on the timing and duration of
treatment.

One possible explanation for the prolonged \emph{C. difficile}
colonization in 5-day PEG treated mice, might be due to the bacteria's
persistence in the mucosal compartment. In fact, it has been
hypothesized that \emph{C. difficile} biofilms may serve as reservoirs
for recurrent infections (35) and \emph{C. difficile} biofilms in the
mucus layer were recently identified in patients as aggregates with
\emph{Fusobacterium nucleatum} (36). There was an interesting pattern of
increased \emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C.
difficile} in both the stool and mucosal communities of PEG-treated mice
suggesting a potential synergy. \emph{Bacteroides} has the potential to
degrade mucus and the osmotic laxative may have allowed
\emph{Bacteroides} to colonize the mucosal niche by degrading mucin
glycans with glycosyl hydrolases that are absent in \emph{C. difficile}
(37). \emph{Bacteroides} persistence in the mucosal tissue might also
have helped \emph{Enterobacteriaceae} to colonize the region, as a
synergy between mucus-degrading \emph{B. fragilis} and \emph{E. coli}
has previously been described (38). A separate study demonstrated
\emph{C. difficile} was present in the outer mucus layer and associated
with \emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using
fluorescent in situ hybridization (FISH) staining (39). However,
protective roles for \emph{Bacteroides} have also been demonstrated. For
example, \emph{B. fragilis} prevented CDI morbidity in a mouse model and
inhibited \emph{C. difficile} adherence \emph{in vitro} (40). In
coculture experiments, \emph{B. longum} decreased \emph{C. difficile}
biofilm formation while \emph{B. thetaiotamicron} enhanced biofilm
formation (41) and \emph{B. dorei} reduced \emph{C. difficile} growth in
a 9-species community \emph{in vitro} (42). Therefore, whether
\emph{Bacteroides} is detrimental or beneficial in the context of
\emph{C. difficile} infection or colonization is still unclear, but the
niche and interactions with other bacteria may contribute.

\emph{Akkermansia} is also a mucin degrader with potentially beneficial
or detrimental roles depending on context in other diseases (43, 44). In
our study, the relative abundance of \emph{Akkermansia} shifted over
time between groups of mice that either cleared \emph{C. difficile} or
had prolonged colonization. In the stool it was initially increased in
mice that cleared \emph{C. difficile}, but shifted after 5-days
post-challenge so that it was increased in mice that had prolonged
colonization. In the context of CDIs, some studies suggest a protective
role (45, 46), while others suggest a detrimental role because
\emph{Akkermansia} was positively correlated with \emph{C. difficile}
(47--50). Because the relative abundance of \emph{Akkermansia} was
dynamic in our study, it is unclear whether \emph{Akkermansia} helps
with clearance of \emph{C. difficile} or allows it to persist. A better
understanding how \emph{C. difficile} interacts with the mucosal
microbiota may lead to insights into CDIs, asymptomatic \emph{C.
difficile} carriage, and colonizatiion resistance.

Despite identifying an altered compositional profile that included
higher relative abundance of the \emph{C. difficile} sequence in the
mucosal tissues of mice treated with 5-day PEG compared to the
clindamycin group, we did not see a difference in histopathology scores
between the groups. One reason there was no difference could be the
\emph{C. difficile} strain used, \emph{C. difficile} 630 results in mild
histopathology summary scores in mice compared to VPI 10463 despite both
strains producing toxin in mice (51). Part of our hypothesis for why
there could have been increased histopathology scores in PEG-treated
mice was because PEG was previously shown to disrupt the mucus layer in
mice. However, recent studies demonstrated that broad spectrum
antibiotics can also disrupt the host mucosal barrier in mice (52, 53).
Further research is needed to tease out the interplay between
medications that influence the mucus layer and different strains of
\emph{C. difficile} in the context of CDIs.

It is more difficult to interpret what are findings mean in the context
of \emph{C. difficile} colonization resistance in human patients. The
main difficulty being that most hospitals recommend not performing
\emph{C. difficile} testing if the patient is currently taking a
laxative. This recommendation is in accordance with the Infectious
Diseases Society of America and Society for Healthcare Epidemiology of
America guidelines (54). The rationale behind the recommendation is that
patients taking laxatives may be asymptomatically colonized with
\emph{C. difficile}, resulting in unnecessary antibiotic treatment
(55--57). Furthermore, some studies identified laxatives as a risk
factor for developing CDIs or recurrent CDIs (58--60) and a recent study
found the proportion of severe CDIs was similar between patients taking
and not taking laxatives (61). However, there have also been some
studies that suggest laxatives are not a risk factor for developing CDIs
(62, 63). Although, it is unclear whether laxatives impact CDI
susceptibility in human paitents, it is clear that laxatives also have a
transient impact on the human microbiota (13, 64--67). Additional
studies to examine the relationship between laxatives, \emph{C.
difficile} colonization, and CDIs are warranted.

Considering laxatives are also used to prepare patients when
administering fecal microbiota transplants via colonoscopy to treat
recurrent CDIs, it will be important to determine whether osmotic
laxatives impact \emph{C. difficile} clearance in the human intestinal
tract. It is still unclear what the best administration route is because
there have been no studies designed to evaluate the best administration
route for FMTs (68). Nevertheless, results from the FMT National
Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
FMTs are highly effective treatment for recurrent CDIs with 90\%
achieving resolution in the 1 month follow-up window (69). A surprising
number of studies continue to hypothesize that PEG or bowel preparation
can clear \emph{C. difficile} spores and toxins despite the paucity of
supporting evidence (20--23). There was even a clinical trial
(NCT01630096) designed to examine whether administering PEG 3350
(NuLYTELY) prior to antibiotic treatment reduced disease severity that
started recruitment in 2012 (70), but no results have been posted to
date. Here we sought to evaluate the impact of treating \emph{C.
difficile} colonized mice with PEG (with or without FMT) and found
clearance was delayed. Further studies are needed to understand the
impact of osmotic laxatives on \emph{C. difficile} colonization
resistance and clearance in human patients receiving FMTs.

@@

\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
25 microliters containing 10\textsuperscript{5} \emph{C. difficile} 630
spores, except for 1 experiment where the concentration was
10\textsuperscript{3} (Fig. 5A). All mock challenged mice received 25 ul
vehicle solution (Ultrapure water). A Dymax stepper pipette was used to
administer the same challenge dose to mice via oral gavage. Mice were
weighed daily throughout the experiment and stool was collected for
quantifying \emph{C. difficile} CFU and 16S rRNA gene sequencing. There
were two groups of mice that received either a PBS or fecal microbiota
transplant (FMT) gavage post-PEG treatment. The fecal microbiota
transplant was prepared with stool samples collected from the mice in
the experiment prior to the start of any treatments. The stool samples
were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS
and glycerol was added to make a 15\% glycerol solution. The solution
was then aliquoted into tubes and stored at -80°C until the day of the
gavage. An aliquot of both the FMT and PBS solutions were also set aside
in the -80°C for 16S rRNA gene sequencing. The day of the gavage,
aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The
supernatant was then transferred to a separate tube to prevent the
gavage needle from clogging with debris during gavage. The PBS solution
that was administered to the other group was also 15\% glycerol. Each
mouse was administered 100 microliters of either the FMT or PBS solution
via gavage. When we refer to mice that cleared \emph{C. difficile}, we
mean that no \emph{C. difficile} was detected in the first serial
dilution (limit of detection: 100 CFU). In some experiments, we
collected tissues for 16SrRNA gene sequencing, histopathology, or both.
For 16S rRNA gene sequencing, we collected small snips of cecum,
proximal colon, and distal colon tissues in microcentrifuge tubes, snap
froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
and colon tissues were placed into separate cassettes, fixed, and then
submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
embedding, and hematoxylin and eosin (H\&E) staining.

@@

\textbf{16S rRNA gene sequencing.} Stool samples were stored at -80°C
and were placed into 96-well plates for DNA extractions and library
preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit
(Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
For library preparation, each plate had a mock community control
(ZymoBIOMICS microbial community DNA standards) and a negative control
(water). The V4 region of the 16S rRNA gene was amplified with the
AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
barcoded primers, as previously described (71). The PCR amplicons were
normalized (SequalPrep normalizatin plate kit from Thermo Fisher
Scientific), pooled and quantified (KAPA library quantification kit from
KAPA Biosystems), and sequenced with the MiSeq system (Illumina).

\textbf{16S rRNA gene sequence analysis.} All sequences were processed
with mothur (v. 1.43) using a previously published protocol (71, 72).
Paired sequencing reads were combined and aligned with the SILVA (v.
132) reference database (73) and taxonomy was assigned with a modified
version of the Ribosomal Database Project reference sequences (v. 16)
(74). The error rate for are sequencing data was 0.0559\% based on the
17 mock communities we ran with the samples. Samples were rarefied to
1,000 sequences, 1,000 times for alpha and beta diversity analyses in
order to account for uneven sequencing across samples. All but 3 of the
17 water controls had fewer than 1000 sequences. PCoAs were generated
based on Bray-Curtis Index distance matrices. Permutational multivariate
analysis of variance (PERMANOVA) tests were performed on
mothur-generated Bray-Curtis distance matrices with the adonis function
from the vegan R package (75).

\textbf{Histopathology.} H\&E stained sections of cecum and colon
tissues collected at either 0, 4, or 6 days post-challenge were coded to
be scored in a blinded manner by a board-certified veterinary
pathologist (ILB). Slides were evaluated using a scoring system
developed for mouse models of \emph{C. difficile} infection (51). Each
slide was evaluated for edema, cellular infiltration, and inflammation
and given a score ranging from 0-4. The summary score was calculated by
combining the scores from the 3 categories and ranged from 0-12.

\textbf{Classification model training and evaluation.} We used the
mikropml package to train and evaluate models to predict \emph{C.
difficile} colonization status 10 days post-challenge where mice were
categorized as either cleared or colonized (76, 77). We removed the
\emph{C. difficile} genus relative abundance data prior to training the
model. Input community relative abundance data at the genus level from 5
days post-challenge was used to generate random forest, logistic
regression, and support vector machine classification models to predict
\emph{C. difficile} colonization status 10 days post-challenge. To
accommodate the small number of samples in our data set we used 50\%
training and 50\% testing splits with repeated 2-fold cross-validation
of the training data for hyperparamter tuning. Permutation importance
was performed as described previously (78) using mikropml (76, 77) with
the random forest model because it had the highest AUROC value.

\textbf{Statistical analysis.} R (v. 4.0.2) and the tidyverse package
(v. 1.3.0) were used for statistical analysis (79, 80). Kruskal-Wallis
tests with Bejamini-Hochberg correction for testing multiple time points
were used to analyze differences in \emph{C. difficile} CFU, mouse
weight change, and alpha diversity between treatment groups. Paired
Wilcoxon rank signed rank tests were used to identify genera impacted by
treatments on matched pairs of samples from 2 time points. Bacterial
relative abundances that varied between treatment groups at the genus
level were identified with the Kruskal-Wallis test with
Benjamini-Hochberg correction for testing all identified OTUs, followed
by pairwise Wilcoxon comparisons with Benjamini-Hocherg correction.

@@

\leavevmode\hypertarget{ref-Frost2021}{}%
35. Frost LR, Cheng JKJ, Unnikrishnan M. 2021. \emph{Clostridioides
difficile} biofilms: A mechanism of persistence in the gut? PLOS
Pathogens 17:e1009348.

\leavevmode\hypertarget{ref-Engevik2021}{}%
36. Engevik MA, Danhof HA, Auchtung J, Endres BT, Ruan W, Bassères E,
Engevik AC, Wu Q, Nicholson M, Luna RA, Garey KW, Crawford SE, Estes MK,
Lux R, Yacyshyn MB, Yacyshyn B, Savidge T, Britton RA, Versalovic J.
2021. Fusobacterium nucleatum adheres to \emph{Clostridioides difficile}
via the RadD adhesin to enhance biofilm formation in intestinal mucus.
Gastroenterology 160:1301--1314.e8.

>>>>

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients were
separated into two groups based on stool consistency, there were similar
microbiota features between samples from CDI patients and \emph{C.
difficile} negative patients with diarrhea compared to non-diarrheal
samples that were \emph{C. difficile} negative (12). The similar
community features between CDI patients and patients with diarrhea
included low alpha diversity and only 6 bacterial taxa with higher
relative abundances in communities from CDI patients. These results led
to the hypothesis that bacterial communities from patients experiencing
diarrhea are susceptible to developing CDIs, regardless of how they
developed diarrhea. For example, laxatives may disrupt colonization
resistance to \emph{C. difficile}.

Depending on the dose administered, osmotic laxatives can lead to
diarrhea and temporarily disrupt the human intestinal microbiota (13).
The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
in Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after the challenge is unclear.

@@

Because there were differences in the mucosal microbiota, including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
histopathology (27). However, we found there was no difference in cecum
and colon scores between clindamycin and PEG-treated mice that were
challenged with \emph{C. difficile} at 4 days post-challenge (Fig. 3D),
the time point typically examined in \emph{C. difficile} 630 challenged
mice (28). We also looked at 6 days post-challenge because that was when
there was a large difference in \emph{C. difficile} density between PEG-
and clindamycin-treated mice (Fig. 1C). Although there was a slight
difference in the histopathology score of the colon between PEG and
clindamycin-treated mice, there was not a significant difference in the
cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
the mucosal microbiota, the impact of \emph{C. difficile} challenge on
the cecum and colon was similar between PEG and clindamycin treated
mice.

@@

While the disruptive effect of antibiotics on \emph{C. difficile}
colonization resistance is well established, the extent to which other
drugs such as laxatives disrupt colonization resistance was unclear. By
following mice treated with an osmotic laxative over time, we found that
a 5-day PEG treatment before challenge resulted in prolonged \emph{C.
difficile} colonization, while a 1-day PEG treatment resulted in
transient colonization without the use of antibiotics. The differences
in \emph{C. difficile} colonization dynamics between the 5- and 1-day
PEG treated mice were associated with differences in the degree to which
treatments disrupted the microbiota. Additionally, the intestinal
communities of 5-day PEG treated mice did not regain colonization
resistance after a 10-day recovery period. In contrast to the other
5-day PEG treatment groups, \emph{C. difficile} was not immediately
detectable in the stools of most of the mice in the 10-day recovery
group. We also examined the impact of PEG treatment after \emph{C.
difficile} challenge. In opposition to the hypothesis suggested by the
literature, we found that PEG treatment prolonged colonization relative
to mice that only received clindamycin treatment. We identified patterns
in the relative abundances of \emph{Bacteroides},
\emph{Enterobacteriaceae}, \emph{Akkermansia},
\emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
associated with prolonged \emph{C. difficile} colonization (Fig. 8).
Overall, our results demonstrated that osmotic laxative treatment alone
rendered mice susceptible to \emph{C. difficile} colonization and the
duration of colonization depended on the length of PEG treatment and
whether treatment was administered before or after challenge.

In addition to altering composition, laxative treatment may alter
microbiota-produced metabolites. A previous study demonstrated that a
5-day treatment of 10\% PEG depleted acetate and butyrate and increased
succinate compared to untreated mice (15). While we did not perform
metabolomic analysis, we did see bacteria known to produce beneficial
metabolites were depleted in mice that cleared \emph{C. difficile}
compared to mice with prolonged colonization (Fig. 7B). For example,
\emph{Oscillibacter valericigenes} can produce the SCFA valerate (30),
and separate studies demonstrated valerate is depleted after clindamycin
treatment and inhibited \emph{C. difficile} growth \emph{in vitro} and
in C57BL/6 mice (31, 32). Similarly, \emph{Acetatifactor} can produce
acetate and butyrate (33), SCFAs that are decreased in mice with
prolonged \emph{C. difficile} infection after antibiotic treatment (34).
Thus protective bacteria and their metabolites could be depleted by
osmotive laxative treatment depending on the timing and duration of
treatment.

One possible explanation for the prolonged \emph{C. difficile}
colonization in 5-day PEG treated mice, might be due to the bacteria's
persistence in the mucosal compartment either within host cells (35) or
together with other bacteria. In fact, it has been hypothesized that
\emph{C. difficile} biofilms may serve as reservoirs for recurrent
infections (36) and \emph{C. difficile} biofilms in the mucus layer were
recently identified in patients as aggregates with \emph{Fusobacterium
nucleatum} (37). There was an interesting pattern of increased
\emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C. difficile}
in both the stool and mucosal communities of PEG-treated mice suggesting
a potential synergy. \emph{Bacteroides} has the potential to degrade
mucus and the osmotic laxative may have allowed \emph{Bacteroides} to
colonize the mucosal niche by degrading mucin glycans with glycosyl
hydrolases that are absent in \emph{C. difficile} (38).
\emph{Bacteroides} persistence in the mucosal tissue might also have
helped \emph{Enterobacteriaceae} to colonize the region, as a synergy
between mucus-degrading \emph{B. fragilis} and \emph{E. coli} has
previously been described (39). A separate study demonstrated \emph{C.
difficile} was present in the outer mucus layer and associated with
\emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using fluorescent in
situ hybridization (FISH) staining (40). However, protective roles for
\emph{Bacteroides} have also been demonstrated. For example, \emph{B.
fragilis} prevented CDI morbidity in a mouse model and inhibited
\emph{C. difficile} adherence \emph{in vitro} (41). In coculture
experiments, \emph{B. longum} decreased \emph{C. difficile} biofilm
formation while \emph{B. thetaiotamicron} enhanced biofilm formation
(42) and \emph{B. dorei} reduced \emph{C. difficile} growth in a
9-species community \emph{in vitro} (43). Therefore, whether
\emph{Bacteroides} is detrimental or beneficial in the context of
\emph{C. difficile} infection or colonization is still unclear, but the
niche and interactions with other bacteria may contribute.

\emph{Akkermansia} is also a mucin degrader with potentially beneficial
or detrimental roles depending on context in other diseases (44, 45). In
our study, the relative abundance of \emph{Akkermansia} shifted over
time between groups of mice that either cleared \emph{C. difficile} or
had prolonged colonization. In the stool it was initially increased in
mice that cleared \emph{C. difficile}, but shifted after 5-days
post-challenge so that it was increased in mice that had prolonged
colonization. In the context of CDIs, some studies suggest a protective
role (46, 47), while others suggest a detrimental role because
\emph{Akkermansia} was positively correlated with \emph{C. difficile}
(48--51). Because the relative abundance of \emph{Akkermansia} was
dynamic in our study, it is unclear whether \emph{Akkermansia} helps
with clearance of \emph{C. difficile} or allows it to persist. A better
understanding how \emph{C. difficile} interacts with the mucosal
microbiota may lead to insights into CDIs, asymptomatic \emph{C.
difficile} carriage, and colonization resistance.

Despite identifying an altered compositional profile that included
higher relative abundance of the \emph{C. difficile} sequence in the
mucosal tissues of mice treated with 5-day PEG compared to the
clindamycin group, we did not see a difference in histopathology scores
between the groups. One reason there was no difference could be the
\emph{C. difficile} strain used, \emph{C. difficile} 630 results in mild
histopathology summary scores in mice compared to VPI 10463 despite both
strains producing toxin in mice (52). Part of our hypothesis for why
there could have been increased histopathology scores in PEG-treated
mice was because PEG was previously shown to disrupt the mucus layer in
mice. However, recent studies demonstrated that broad spectrum
antibiotics can also disrupt the host mucosal barrier in mice (53, 54).
Further research is needed to tease out the interplay between
medications that influence the mucus layer and different strains of
\emph{C. difficile} in the context of CDIs.

The extent to which laxatives disrupt \emph{C. difficile} colonization
resistance in human patients is unclear based on the current literature.
The main difficulty being that most hospitals recommend not performing
\emph{C. difficile} testing if the patient is currently taking a
laxative. This recommendation is in accordance with the Infectious
Diseases Society of America and Society for Healthcare Epidemiology of
America guidelines (55). The rationale behind the recommendation is that
patients taking laxatives may be asymptomatically colonized with
\emph{C. difficile}, resulting in unnecessary antibiotic treatment
(56--58). Furthermore, some studies identified laxatives as a risk
factor for developing CDIs or recurrent CDIs (59--61) and a recent study
found the proportion of severe CDIs was similar between patients taking
and not taking laxatives (62). However, there have also been some
studies that suggest laxatives are not a risk factor for developing CDIs
(63, 64). Although, it is unclear whether laxatives impact CDI
susceptibility in human paitents, it is clear that laxatives also have a
transient impact on the human microbiota (13, 65--68). Additional
studies to examine the relationship between laxatives, \emph{C.
difficile} colonization, and CDIs are warranted.

Considering laxatives are also used to prepare patients when
administering fecal microbiota transplants via colonoscopy to treat
recurrent CDIs, it will be important to determine whether osmotic
laxatives impact \emph{C. difficile} clearance in the human intestinal
tract. It is still unclear what the best administration route is because
there have been no studies designed to evaluate the best administration
route for FMTs (69). Nevertheless, results from the FMT National
Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
FMTs are highly effective treatment for recurrent CDIs with 90\%
achieving resolution in the 1 month follow-up window (70). A surprising
number of studies continue to hypothesize that PEG or bowel preparation
can clear \emph{C. difficile} spores and toxins despite the paucity of
supporting evidence (20--23). There was even a clinical trial
(NCT01630096) designed to examine whether administering PEG 3350
(NuLYTELY) prior to antibiotic treatment reduced disease severity that
started recruitment in 2012 (71), but no results have been posted to
date. Here we sought to evaluate the impact of treating \emph{C.
difficile} colonized mice with PEG (with or without FMT) and found
clearance was delayed. Further studies are needed to understand the
impact of osmotic laxatives on \emph{C. difficile} colonization
resistance and clearance in human patients receiving FMTs.

@@

\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
25 microliters containing 10\textsuperscript{5} \emph{C. difficile} 630
spores, except for 1 experiment where the concentration was
10\textsuperscript{3} (Fig. 5A). All mock challenged mice received 25 ul
vehicle solution (Ultrapure water). A Dymax stepper pipette was used to
administer the same challenge dose to mice via oral gavage. Mice were
weighed daily throughout the experiment and stool was collected for
quantifying \emph{C. difficile} CFU and 16S rRNA gene sequencing. Fresh
stool samples were collected from each mouse and split into two separate
tubes. One tube was transferred to an anaerobic chamber the same day the
sample was collected to quantify \emph{C. difficile}, while the other
tube was snap frozen in liquid nitrogen and stored in the -80°C for 16S
rRNA sequencing. There were two groups of mice that received either a
PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The
fecal microbiota transplant was prepared with stool samples collected
from the mice in the experiment prior to the start of any treatments.
The stool samples were transferred to an anaerobic chamber and diluted
1:10 in reduced PBS and glycerol was added to make a 15\% glycerol
solution. The solution was then aliquoted into tubes and stored at -80°C
until the day of the gavage. An aliquot of both the FMT and PBS
solutions were also set aside in the -80°C for 16S rRNA gene sequencing.
The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM
for 1 minute. The supernatant was then transferred to a separate tube to
prevent the gavage needle from clogging with debris during gavage. The
PBS solution that was administered to the other group was also 15\%
glycerol. Each mouse was administered 100 microliters of either the FMT
or PBS solution via gavage. When we refer to mice that cleared \emph{C.
difficile}, we mean that no \emph{C. difficile} was detected in the
first serial dilution (limit of detection: 100 CFU). In some
experiments, we collected tissues for 16SrRNA gene sequencing,
histopathology, or both. For 16S rRNA gene sequencing, we collected
small snips of cecum, proximal colon, and distal colon tissues in
microcentrifuge tubes, snap froze in liquid nitrogen, and stored at
-80°C. For histopathology, cecum and colon tissues were placed into
separate cassettes, fixed, and then submitted to McClinchey Histology
Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and
eosin (H\&E) staining.

@@

\textbf{16S rRNA gene sequencing.} Stool samples were stored at -80°C
and were placed into 96-well plates for DNA extractions and library
preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit
(Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
For library preparation, each plate had a mock community control
(ZymoBIOMICS microbial community DNA standards) and a negative control
(water). The V4 region of the 16S rRNA gene was amplified with the
AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
barcoded primers, as previously described (72). The PCR amplicons were
normalized (SequalPrep normalizatin plate kit from Thermo Fisher
Scientific), pooled and quantified (KAPA library quantification kit from
KAPA Biosystems), and sequenced with the MiSeq system (Illumina).

\textbf{16S rRNA gene sequence analysis.} All sequences were processed
with mothur (v. 1.43) using a previously published protocol (72, 73).
Paired sequencing reads were combined and aligned with the SILVA (v.
132) reference database (74) and taxonomy was assigned with a modified
version of the Ribosomal Database Project reference sequences (v. 16)
(75). The error rate for are sequencing data was 0.0559\% based on the
17 mock communities we ran with the samples. Samples were rarefied to
1,000 sequences, 1,000 times for alpha and beta diversity analyses in
order to account for uneven sequencing across samples. All but 3 of the
17 water controls had fewer than 1000 sequences. PCoAs were generated
based on Bray-Curtis Index distance matrices. Permutational multivariate
analysis of variance (PERMANOVA) tests were performed on
mothur-generated Bray-Curtis distance matrices with the adonis function
from the vegan R package (76).

\textbf{Histopathology.} H\&E stained sections of cecum and colon
tissues collected at either 0, 4, or 6 days post-challenge were coded to
be scored in a blinded manner by a board-certified veterinary
pathologist (ILB). Slides were evaluated using a scoring system
developed for mouse models of \emph{C. difficile} infection (52). Each
slide was evaluated for edema, cellular infiltration, and inflammation
and given a score ranging from 0-4. The summary score was calculated by
combining the scores from the 3 categories and ranged from 0-12.

\textbf{Classification model training and evaluation.} We used the
mikropml package to train and evaluate models to predict \emph{C.
difficile} colonization status 10 days post-challenge where mice were
categorized as either cleared or colonized (77, 78). We removed the
\emph{C. difficile} genus relative abundance data prior to training the
model. Input community relative abundance data at the genus level from 5
days post-challenge was used to generate random forest, logistic
regression, and support vector machine classification models to predict
\emph{C. difficile} colonization status 10 days post-challenge. To
accommodate the small number of samples in our data set we used 50\%
training and 50\% testing splits with repeated 2-fold cross-validation
of the training data for hyperparamter tuning. Permutation importance
was performed as described previously (79) using mikropml (77, 78) with
the random forest model because it had the highest AUROC value.

\textbf{Statistical analysis.} R (v. 4.0.2) and the tidyverse package
(v. 1.3.0) were used for statistical analysis (80, 81). Kruskal-Wallis
tests with Bejamini-Hochberg correction for testing multiple time points
were used to analyze differences in \emph{C. difficile} CFU, mouse
weight change, and alpha diversity between treatment groups. Paired
Wilcoxon rank signed rank tests were used to identify genera impacted by
treatments on matched pairs of samples from 2 time points. Bacterial
relative abundances that varied between treatment groups at the genus
level were identified with the Kruskal-Wallis test with
Benjamini-Hochberg correction for testing all identified OTUs, followed
by pairwise Wilcoxon comparisons with Benjamini-Hocherg correction.

@@

\leavevmode\hypertarget{ref-CastroCrdova2021}{}%
35. Castro-Cordova P, Mora-Uribe P, Reyes-Ramirez R, Cofre-Araneda G,
Orozco-Aguilar J, Brito-Silva C, Mendoza-Leon MJ, Kuehne SA, Minton NP,
Pizarro-Guajardo M, Paredes-Sabja D. 2021. Entry of spores into
intestinal epithelial cells contributes to recurrence of
\emph{Clostridioides difficile} infection. Nature Communications 12.

\leavevmode\hypertarget{ref-Frost2021}{}%
36. Frost LR, Cheng JKJ, Unnikrishnan M. 2021. \emph{Clostridioides
difficile} biofilms: A mechanism of persistence in the gut? PLOS
Pathogens 17:e1009348.

\leavevmode\hypertarget{ref-Engevik2021}{}%
37. Engevik MA, Danhof HA, Auchtung J, Endres BT, Ruan W, Bassères E,
Engevik AC, Wu Q, Nicholson M, Luna RA, Garey KW, Crawford SE, Estes MK,
Lux R, Yacyshyn MB, Yacyshyn B, Savidge T, Britton RA, Versalovic J.
2021. Fusobacterium nucleatum adheres to \emph{Clostridioides difficile}
via the RadD adhesin to enhance biofilm formation in intestinal mucus.
Gastroenterology 160:1301--1314.e8.
